Non-parametric approaches to optimal policy are more robust

### 2013-04-08 inst/doc/nonparametric-bayes.tex
writing writing -- draft sent to Marc & Steve

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological
interactions involved, in practice real mechanisms are often unknown and
the true dynamics too complex to be captured by simple models
{[}@refs{]}. While simple mechanistic models can nevertheless provide
imporant insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100\% effective to eliminate the
transmission of a virus -- such approaches are not well-suited for use
in forecasting outcomes of potential management options. Non-parametric
approaches offer a more flexible alternative that can both more
accurately reflect the data available while also representing greater
uncertainty in areas (of state-space) where data is lacking. Ecological
research and management strategy should pay closer attention to the
opportunities and challenges nonparametric modeling can offer.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown {[}@refs{]}. Despite this,
quantitative policy optimization in these systems has almost exclusively
been based on simple, mechanistically motivated parametric models. There
are several reasons for this. First, limited available data almost
always percludes any accurate estimate of more complicated models.
Second, the computational demands of determining an optimal stategy in a
sequential decision-theory problem suffer from the ``curse of
dimensionality'' that makes it difficult to consider more complicated
models, states, or actions. Unfortunately, these challenges also make it
particularly difficult to reflect the true uncertainty either in the
parameters of the given models (parametric uncertainty) or in the
structure of the models themselves (structural uncertainty). Further
unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue. Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties of
solving such a system. Cressie et al. (2009) provides a good account of
the successess and challenges of the approach. Applying these approaches
in the management context of sequential decision making, in which
forecasts must be obtained over a range of possible actions and updated
regularly as new information arrives makes such an approach less
feasible still.

@@

 \textbf{This part is overstated and unnecessary} Though machine
learning approaches have begun to appear in the ecological and
conservation literature (e.g.~in species distribution models), including
the Gaussian process based approach used here (Munch et al. 2005),
overall machine learning remains an unfamiliar and somewhat untrusted
approach for many ecologists. One potential barrier to their adoption is
the absence of a framework for applying machine learning approaches to
resource management problems.

 transition sentences!?

@@

 Okay, perhaps that's novel, but it's pretty trivial. Isn't it obvious
to everyon that it's trivial?

 paraphrase advantages of machine learning

@@

A crux of this approach is correctly specifying the functional form of
$f$, along with its parameters. The standard approach {[}@refs{]} uses
one of a handful of common parametric models representing the
stock-recruitment relationship, usually after estimating the model
parameters from any available existing data. Though modifications of
this approach mentioned in the introduction can permit additional
sources of uncertainty such as measurement error in the stock
assessment, implementation errors in the harvest policy, (Clark and
Kirkwood 1986; Roughgarden and Smith 1996; Sethi et al. 2005)
uncertainty in parameters (Mangel and Clark 1985; Schapaugh and Tyre
2013) or model structure (B. K. Williams 2001; Athanassoglou and
Xepapadeas 2012), we focus on the simplest example to illustrate the
comparison, where only a single form of uncertainty (the growth shocks
$Z_t$) is present in the underlying model, making additional modes
unnecessary. Add comment / apology on single species dynamics

@@

The essense of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.\\
Figure 1 include curves drawn from the posterior density?

 Do we need more specifics on Gaussian process as an approximation to
parametric models? Discussion of Gaussian process vs other machine
learning / forecasting approaches that have less clear statistical
foundations? If so, does this all belong in the discussion?

@@

The posterior Gaussian process is completely defined by the expected
value and covariance matrix at a defined set of training points. For
simplicty we will consider a these points to fall on a discrete, uniform
grid $x$ of 101 points from 0 to 15 (1.5 times the positive equilibrium
$K$). Again to keep things simple we will use this same grid
discritization for the parametric approach. Other options for choosing
the grid points, including collocation methods and functional basis
expansion (or even using Guassian processes in place of the discrete
optimization; an entirely different context in which GP can be used in
SDP, see (Deisenroth, Rasmussen, and Peters 2009)) could also be
considered.

The transition matrix $\mathbf{F}$ is thus an 101 by 101 matrix for
which the ${i,j}$ entry gives the probability of transitioning into
state $x_i$ given that the system is in state $x_j$ in the previous
timestep. To generate the transition matrix based on the posterior GP,
we need only the expected values at each grid point and the
corresponding variances (the diagonal of the covariance matrix), as
shown in Figure 1. Given the mean at each gridpoint as the length 101
vector $E$ and variance $V$, the probability of transitioning from state
$x_i$ to state $x_j$ is simply
$\mathcal{N}\left(x_j \vert  \mu = E_i, \sigma = \sqrt{V_i}\right)$,
where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix.

@@

@@

@@

\subsubsection{The optimal control problem}

\subsubsection{Discussion on how we compare performance of policies}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Replicate stochastic simulations
\item
  Sensitivity analysis (Figure 4).
\end{itemize}

@@

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). To compare
the performance of nonparametric and parametric approaches in an example
that is easy to conceptualize, we will focus on a simple parametric
model for a single species (derived from fist principles by Allen et al.
2005) as our underlying ``reality''.

@@

\subsection{Sample training data}

@@

Should we include any emprical examples? Using data simulated from a
specified model rather than empirical data permits the comparison
against the true underlying dynamics, setting a bar for the optimal
performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

\subsubsection{Discussion of maximum likelihood estimated models}

(MLE models will assume the noise is log-normal, which it is in the
simulation).

Which estimates a Ricker model with $r =$ 1.8501, $K =$ 9.8091, and the
Allen Allele model with $r =$ 2.8079, $K =$ 11.8235 and $C =$ 7.2159.

\subsection{Figure 1:}

\emph{Shows the inferred Gaussian Process compared to the true and
parametric models. Refer to the appendix for details on the GP
posteriors, etc.}

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. (The GP is conditioned on
0,0, shown as a pseudo-data point).}
\end{figure}

\subsection{Figure 2:}

\emph{The take-home message, showing that the GP is closest to the
optimal strategy, while the parametric methods are less accurate.
Visualizing the policy may be more useful for the technical reader, the
general audience may prefer Figure 3 showing all replicates of the
population collapse under the parametric model and not under the GP.}

\begin{figure}[htbp]
\centering
\includegraphics{figure/policies_plot.pdf}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

\subsection{Figure 3:}

\emph{Figure 3 is a less abstract and more visceral visualization of the
take-home message, with the structurally inaccurate model leading
universally to a collapse of the fishery and very few profits, while the
Gaussian process performs nearly optimally. The parametric approach even
with the correct underlying structure does not perform optimally,
choosing in this case to under-fish (may need to show harvest dynamics
since that is not clear from the figure! Also isn't general, sometimes
does optimally, sometimes over-fishes. Perhaps need to show more
examples.) May need to show profits too?}

\begin{figure}[htbp]
\centering
\includegraphics{figure/sim_plot.pdf}
\caption{Gaussian process inference outperforms parametric estimates.
Shown are 100 replicate simulations of the stock dynamics (eq 1) under
the policies derived from each of the estimated models, as well as the
policy based on the exact underlying model.}
\end{figure}

\subsection{Figure 4:}

\emph{Shows the sensitivity analysis. A histogram of distribution of
yield over stochastic realizations, showing that the qualitative results
do not depend on the stochastic realization of the training data here,
or on the parameters of the underlying model, though quantitative
differences are visible.}

\section{Discussion / Conclusion}

Non-parametric methods have received far too little attention in
ecological modeling efforts that are aimed at improved conservation
planning and decision making support.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Adapting a non-parametric approach requires modification of existing
  methods for decision theory. We have illustrated how this might be
  done in the context of stochastic dynamic programming, opening the
  door for substantial further research into how these applications
  might be improved.
\end{itemize}

Three elements contribute to the machine learning model performing as
well or better than the mechanisitic models in this context.

\subsubsection{1. Relevant state space}

The dynamics occur over a range of state-space in which the machine
learning model closely matches the predictions of y implausible model
performs as well or better than the more plausible model.

This is the most obvious and immediate reason why the shortcomings that
appear to make the model implausible do not make it useless. This effect
is enhanced in our example because the management actions help drive the
system towards rather than away from this region of state-space.

\subsubsection{2. Predictive accuracy}

This application relies only on the predictive accuracy of the model,
not an interpretation of the parameters.

 wow, run-on tangent? Predictive accuracy is not the goal of all
modeling, as ecologists have been observing for as long as they made
models (perhaps none more memorably than Levins (1969)). Mechanistic
modeling is at its most powerful not when it is used to directly
forecast future states but when it provides an understanding of how to
approach a problem. SIR-type models from the epidemiological literature
are a good example. While the simplest SIR models have little predictive
power over the outbreak intensity or timing at a particular location,
they provide a powerful understanding of the spread of an infection in
terms of a single, biologically meaningful parameter: $R_0$, the basic
reproductive number. From the model, it becomes clear that management
need not vaccinate every member of the population to stop the spread,
but rather it suffices to vaccinate a sufficient fraction of the
population to reduce $R_0$ below 1.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Better express ambiguity in regions where data is not available
\end{enumerate}

One of the greatest strengths of mechanistic models is their greatest
weakness as well.

@@

 Jargony and vague While we have highlighted certain generic of this
problem that allow the nonparametric approach to excel -- short
timescales between successive observations and actions, the accuracy in
the appropriate region of state space, the ability to express
uncertainty outside the observed data -- there are equally several
aspects for which the nonparametric approach is at a relative
disadvantage.

 Model complexity In this simulated example, the underlying dynamics are
truly governed by a simple parametric model, allowing the parametric
approaches to be more accurate. Similarly, because the dynamics are
one-dimensional dynamics and lead to stable nodes (rather than other
attractors such as limit-cycles resulting in oscillations), the training
data provides relatively limited information about the dynamics. For
these reasons, we anticipate that in higher-dimensional examples
characteristic of ecosystem management problems that the machine
learning approach will prove even more valuable.

 Data complexity. Perhaps too far out of scope\ldots{} The machine
learning approach is also better suited to complex and disparate data.
Incorporating various sources of information into mechanistic models can
be an immensely difficult due to the increased complexity involved. Only
thanks to tandem advances in increasing computational power and
hierarchical statistical methodology have we been able to tackle such
intuitively important complexity (and the potentially new available data
that accompanies it) such as spatial distribution, heterogeneities of
space, time, and individuals, to shift to ecosystem-based approaches
from single-species based approaches. Without the need to formulate
mechanisms, many modern machine learning algorithms can leverage
potential information from all available sources of data directly. The
algorithms can recognize unanticipated or subtle patterns in large data
sets that enable more accurate predictions than mechanistic models that
are formulated at a more macroscopic level.

 * Discuss constant escapement in model, in policies.

\begin{itemize}
\item
  Limitations of this comparison: Are the maximum-likelihood solutions a
  straw man?
\item
  Discussion of alternative related approaches: POMDP/MOMDP 
\end{itemize}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Multiple species
\item
  Online learning
\item
  Multiple step-ahead predictions
\item
  Explicitly accomidating additional uncertainties
\item
  Improving inference of optimal policy from the GP
\end{itemize}

@@

@@

\subsubsection{The Guassian process}

@@

\begin{longtable}[c]{@{}ll@{}}
\hline\noalign{\medskip}
parameter & interpretation
\\\noalign{\medskip}
\hline\noalign{\medskip}
$\sigma^2$ & The process noise (from the kernel)
\\\noalign{\medskip}
$\tau^2$ & Variance around the mean
\\\noalign{\medskip}
$\beta$\_0 & The mean is given by a linear model of slope $\beta$
\\\noalign{\medskip}
$d$ & The length-scale of the covariance function
\\\noalign{\medskip}
$n$ & The observation error
\\\noalign{\medskip}
\hline
\noalign{\medskip}
\caption{Table of parameters for the Gaussian process}
\end{longtable}

@@

The posterior Gaussian process is completely defined by the expected
value and covariance matrix at a defined set of training points. For
simplicty we will consider a these points to fall on a discrete, uniform
grid $x$ of 101 points from 0 to 15 (1.5 times the positive equilibrium
$K$). Again to keep things simple we will use this same grid
discritization for the parametric approach. Other options for choosing
the grid points, including collocation methods and functional basis
expansion (or even using Guassian processes in place of the discrete
optimization; an entirely different context in which GP can be used in
SDP, see (Deisenroth, Rasmussen, and Peters 2009)) could also be
considered.

The transition matrix $\mathbf{F}$ is thus an 101 by 101 matrix for
which the ${i,j}$ entry gives the probability of transitioning into
state $x_i$ given that the system is in state $x_j$ in the previous
timestep. To generate the transition matrix based on the posterior GP,
we need only the expected values at each grid point and the
corresponding variances (the diagonal of the covariance matrix), as
shown in Figure 1. Given the mean at each gridpoint as the length 101
vector $E$ and variance $V$, the probability of transitioning from state
$x_i$ to state $x_j$ is simply
$\mathcal{N}\left(x_j \vert  \mu = E_i, \sigma = \sqrt{V_i}\right)$,
where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/posteriors.pdf}
\caption{Histogram of posterior distributions for the estimated Gaussian
Process shown in Figure 1. Prior distributions overlaid.}
\end{figure}

@@

@@

Levins, Richard. 1969. ``Some demographic and genetic consequences of
environmental heterogeneity for biological control.'' \emph{Bulletin of
the Entomological Society of America} 15 (3): 237--240.
\url{http://www.ingentaconnect.com/content/esa/besa/1969/00000015/00000003/art00009}.

>>>>

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological
interactions involved, in practice real mechanisms are often unknown and
the true dynamics too complex to be captured by simple models
{[}@refs{]}. While simple mechanistic models can nevertheless provide
important insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100\% effective to eliminate the
transmission of a virus -- such approaches are not well-suited for use
in forecasting outcomes of potential management options. Non-parametric
approaches offer a more flexible alternative that can both more
accurately reflect the data available while also representing greater
uncertainty in areas (of state-space) where data is lacking. Ecological
research and management strategy should pay closer attention to the
opportunities and challenges nonparametric modeling can offer.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown {[}@refs{]}. Despite this,
quantitative policy optimization in these systems has almost exclusively
been based on simple, mechanistically motivated parametric models. There
are several reasons for this. First, limited available data almost
always precludes any accurate estimate of more complicated models.
Second, the computational demands of determining an optimal strategy in
a sequential decision-theory problem suffer from the ``curse of
dimensionality'' that makes it difficult to consider more complicated
models, states, or actions. Unfortunately, these challenges also make it
particularly difficult to reflect the true uncertainty either in the
parameters of the given models (parametric uncertainty) or in the
structure of the models themselves (structural uncertainty). Further
unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue. Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties of
solving such a system. Cressie et al. (2009) provides a good account of
the successes and challenges of the approach. Applying these approaches
in the management context of sequential decision making, in which
forecasts must be obtained over a range of possible actions and updated
regularly as new information arrives makes such an approach less
feasible still.

@@

 \textbf{\emph{This part is overstated and unnecessary}} Though machine
learning approaches have begun to appear in the ecological and
conservation literature (e.g.~in species distribution models), including
the Gaussian process based approach used here (Munch et al. 2005),
overall machine learning remains an unfamiliar and somewhat untrusted
approach for many ecologists. One potential barrier to their adoption is
the absence of a framework for applying machine learning approaches to
resource management problems.

\textbf{\emph{transition sentences!?}}

@@

\textbf{\emph{Okay, perhaps that's novel, but it's pretty trivial. Isn't
it obvious to everyone that it's trivial?}}

\textbf{\emph{paraphrase advantages of machine learning}}

@@

A crux of this approach is correctly specifying the functional form of
$f$, along with its parameters. The standard approach {[}@refs{]} uses
one of a handful of common parametric models representing the
stock-recruitment relationship, usually after estimating the model
parameters from any available existing data. Though modifications of
this approach mentioned in the introduction can permit additional
sources of uncertainty such as measurement error in the stock
assessment, implementation errors in the harvest policy, (Clark and
Kirkwood 1986; Roughgarden and Smith 1996; Sethi et al. 2005)
uncertainty in parameters (Mangel and Clark 1985; Schapaugh and Tyre
2013) or model structure (B. K. Williams 2001; Athanassoglou and
Xepapadeas 2012), we focus on the simplest example to illustrate the
comparison, where only a single form of uncertainty (the growth shocks
$Z_t$) is present in the underlying model, making additional modes
unnecessary. \textbf{\emph{Add comment / apology on single species
dynamics?}}

@@

The essence of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.\\
\textbf{\emph{Figure 1 include curves drawn from the posterior
density?}}

\textbf{\emph{Do we need more specifics on Gaussian process as an
approximation to parametric models? Discussion of Gaussian process vs
other machine learning / forecasting approaches that have less clear
statistical foundations? If so, does this all belong in the
discussion?}}

@@

Once the posterior Gaussian process (GP) has been estimated (e.g.~see
Munch et al. 2005), it is necessary to adapt it in place of the
parametric equation for the stochastic dynamic programming (SDP)
solution (see Mangel and Clark 1988 for a detailed description of
parametric SDP methods) to the optimal policy. The essence of the idea
is straight forward -- we will use the estimated GP in place of the
parametric growth function to determine the stochastic transition matrix
on which the SDP calculations are based. The SDP is solved in a
discretized state space -- both the continuously valued population
densities $X$ and harvest quotas $h$ are first mapped to a bounded,
discrete grid. (For simplicity we will consider a uniform grid, though
for either parametric or GP-based SDP it is often advantageous to use a
non-uniform discretization such as a basis function representation,
e.g.~see (Deisenroth, Rasmussen, and Peters 2009)).

The SDP approach then computes a transition matrix, $\mathbf{F}$. We
demonstrate that calculation is just as straight forward based on the GP
as it is in the classical context using the parametric model. The
${i,j}$ of the transition matrix $F$ entry gives the probability of
transitioning into state $x_i$ given that the system is in state $x_j$
in the previous time-step. To generate the transition matrix based on
the posterior GP, we need only the expected values at each grid point
and the corresponding variances (the diagonal of the covariance matrix),
as shown in Figure 1. Given the mean of the GP posterior at each
grid-point as the vector $E$ and variance at that point as vector $V$,
the probability of transitioning from state $x_i$ to state $x_j$ is

\[\mathcal{N}\left(x_j \vert  \mu = E_i, \sigma = \sqrt{V_i}\right)\]

where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix. The
process can then be repeated for each possible discrete value of our
control variable, (harvest $h$).

\textbf{Pseudocode for the determining the transition matrix from the
GP}

@@

Using the discrete transition matrix we may write down the Bellman
recursion defining the stochastic dynamic programming iteration:

@@

where $V(x_t)$ is the value of being at state $x$ at time $t$, $h$ is
control (harvest level) chosen. Numerically, the maximization is
accomplished as follows. Consider the set of possible control values to
be the discrete values corresponding the grid of stock sizes. Then for
each $h_t$ there is a corresponding transition matrix $\mathbf{F}_h$
determined as described above but with mean $\mu = x_j - h_t$. Let
$\vec{V_t}$ be the vector whose $i$th element corresponds to the value
of having stock $x_i$ at time $t$. Then let $\Pi_h$ be the vector whose
$i$th element indicates the profit from harvesting at intensity $h_t$
given a population $x_i$ (e.g. $\max(x_i, h_t)$ since one cannot harvest
more fish then the current population size). Then the Bellman recursion
can be given in matrix form as

@@

This completes the algorithm adapting the GP to the sequential
decision-making problem through SDP, which we believe has not yet been
demonstrated in the literature. We provide an R package implementation
of this, along with the Gaussian process inference, in the supplemental
materials.

@@

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see Hughes et al. (2013)), management is most often concerned with
avoiding potentially catostrophic tipping points before any data is
available at or following a transition that would more clearly reveal
these regime shift dynamics (e.g. Bestelmeyer et al. 2012).

To illustrate the value of the non-parametric Bayesian approach to
management, we focus on example of a system containing such a tipping
point whose dynamics can still be described by a simple, one-dimensional
parametric model.\\We will focus on a simple parametric model for a
single species (derived from fist principles by Allen et al. 2005) as
our underlying ``reality''.

@@

@@

\textbf{\emph{Should we include any emprical examples?}}

\subsubsection{Estimating parametric models}

The maximum likelihood estimates (MLE) will assume that the process
noise $Z_t$ is log-normal, as in the simulation itself. In reality the
noise structure might differ from this assumption, making this a
best-case scenario for the MLE approach.\\The GP must also make an
assumption about the noise structure -- in this case we assume additive
noise, though multiplicative noise would be more accurate. This will
illustrate the flexibility of the GP in fitting different noise
structures even without knowledge of the correct assumptions.

The GP estimate is Bayesian approach which requires specification of
prior distributions for the hyper-parameters (see Appendix for details,
or @Rassmussen1996, Munch et al. 2005 for background), as given in Table
1. Then given the data, both the GP and the MLE parametric models can be
estimated as described above, as we implement in R in the code provided
in supplementary materials.

@@

Figure 1 shows the inferred Gaussian Process compared to the true and
the MLE parametric models, predicting the relationship between observed
stock size (x-axis) to the stock size following recruitment the
following year. The gray shading shows the uncertainty estimated by the
GP. Unlike the MLE estimates, the GP uncertainty changes depending on
where the data is available, allowing for a much greater level of
uncertainty at low stock sizes. While it would be natural (and straight
forward) to condition the GP on passing through the origin (0,0) (see
appendix), the estimate shown here is based only on the observed data.
The observed data from which each model is estimated is also shown. The
observations come from only a limited region of state space
corresponding to unharvested or weakly harvested system. No observations
occur at the theoretical optimum (7.5 units, see Figure 2) or near the
tipping point (5 units).

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.png}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance (grey shading)
is also shown. The training data is also shown as black points.}
\end{figure}

The resulting optimal management strategy based on each of the inferred
models is shown in Figure 2, against the optimal strategy given the true
underlying dynamics. Policies are shown in terms of target escapement,
$S_t$. Under models such as this a constant escapement policy is
expected to be optimal (Reed 1979), whereby population levels below a
certain size $S$ are unharvested, while above that size the harvest
strategy aims to return the population to $S$, resulting in the
hockey-stick shaped policies shown. Only the GP policy deviates slightly
from this. The MLE parametric model corresponding to parametric
uncertainty alone (gold line labeled ``Parametric'', that is, using the
same structural form as the underlying simulation) nevertheless
overestimates the carrying capacity of the stock and therefore
underestimates the harvest rates, always waiting for the stock to rise
further than is actually possible. The Ricker model (green line, showing
``Structural'' inaccuracies) does not foresee the tipping point dynamics
reflected in the concavity of the other three models (Figure 1), and
selects a much more aggressive harvesting strategy.

\begin{figure}[htbp]
\centering
\includegraphics{figure/policies_plot.png}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model.}
\end{figure}

The consequences of managing 100 replicate realizations of the simulated
fishery under each of the policies estimated is shown in Figure 3. As
expected from the policy curves, the structurally correct model
(``Parametric'' uncertainty, gold) underharvests, leaving the stock to
vary around it's unfished optimum. The GP overharvests the population
slightly relative to the optimal value. The structurally incorrect
Ricker model (``Structural'', green line) overharvests the population
passed the tipping point consistently, resulting in the immediate crash
of the stock and thus derives minimal profits.

\textbf{\emph{(may need to show harvest dynamics since that is not clear
from the figure! Also isn't general, sometimes does optimally, sometimes
over-fishes. Perhaps need to show more examples.) May need to show
profits too?}}

\begin{figure}[htbp]
\centering
\includegraphics{figure/sim_plot.png}
\caption{Gaussian process inference outperforms parametric estimates.
Shown are 100 replicate simulations of the stock dynamics (eq 1) under
the policies derived from each of the estimated models, as well as the
policy based on the exact underlying model.}
\end{figure}

The results shown in Figures 1-3 are not unique to the simulated data or
models chosen here, but arises across a range of parameter values and
simulations as shown in the supplemental figures. The results across
this range can most easily be compared by the relative differences in
net present value realized by each of the approaches, as shown in Figure
4. The Gaussian Process most consistently realizes a value close to the
optimal solution, and importantly avoids ever driving the system across
the tipping point, which results in the near-zero value cases in the
parametric models.

\begin{figure}[htbp]
\centering
\includegraphics{figure/figure4.png}
\caption{Histograms of the realized net present value of the fishery
over a range of simulated data and resulting parameter estimates. For
each data set, the three models are estimated as described above. Values
plotted are the averages of a given policy over 100 replicate
simulations. Details and code provided in the supplement.}
\end{figure}

\section{Discussion}

Non-parametric Bayesian methods have received far too little attention
in ecological modeling efforts that are aimed at improved conservation
planning and decision making support. Such approaches may be
particularly useful when the available data is restricted to a limited
area of state-space, which can lead parametric models to underestimate
the uncertainty in dynamics at population levels (states) which have not
been observed. One reason for the relative absence of nonparametric
approaches in the natural resource management context may be the lack of
existing approaches for adapting the non-parametric Bayesian models
previously proposed (Munch et al. 2005) to a decision-theoretic
framework. Adapting a non-parametric approach requires modification of
existing methods for decision theory. We have illustrated how this might
be done for a classic stochastic dynamic programming problem, opening
the door for substantial further research into how these applications
might be improved.

While non-parametric Bayesian approaches will not always be preferable
to simple mechanistic models, we highlight three aspects of the problem
consider here that make these methods particularly valuable. These
aspects are common to many conservation decision making problems, which
thus merit greater use of non-parametric machine learning approaches
that can best take advantage of them.

\subsubsection{1. Large uncertainty where the data is poor}

The parametric models perform worst when they propose a management
strategy outside the range of the observed data. The non-parametric
Bayesian approach, in contrast, allows a predictive model that expresses
a great deal of uncertainty about the probable dynamics outside the
observed range, while retaining very good predictive accuracy in the
range observed. The management policy dictated by the GP balance this
uncertainty against the immediate value of the harvest, and act to
stabilize the population dynamics in a region of state space in which
the predictions can be reliably reflected by the data.

\subsubsection{2. Predictive accuracy where data is good}

While expressing larger uncertainty outside the observed data, the GP
can also provide a better fit with smaller uncertainty inside the range
of the observed data. This arises from the greater flexibility of the
Gaussian process, which describes a large family of possible
curves.\\While in a parametric context this over-fitting would be more
worrisome -- a high-degree polynomial could fit the data even better --
those concerns are driven by the resulting parametric fit outside the
data, which may involve wild oscillations unsupported by the data. As we
have seen in \#1, the GP is less vulnerable to such unjustified
predictions outside the data, and is meanwhile free to benefit from the
greater fit where the data is available.

@@

@@

 Jargony and vague While we have highlighted certain generic of this
problem that allow the nonparametric approach to excel -- short
timescales between successive observations and actions, the accuracy in
the appropriate region of state space, the ability to express
uncertainty outside the observed data -- there are equally several
aspects for which the nonparametric Bayesian approach is nevertheless at
a relative disadvantage. In this simulated example, the underlying
dynamics are truly governed by a simple parametric model, allowing the
parametric approaches to be more accurate. Similarly, because the
dynamics are one-dimensional dynamics and lead to stable nodes (rather
than other attractors such as limit-cycles resulting in oscillations),
the training data provides relatively limited information about the
dynamics. For these reasons, we anticipate that in higher-dimensional
examples characteristic of ecosystem management problems that the
machine learning approach will prove even more valuable.

 Data complexity. Perhaps too far out of scope\ldots{} The nonparametric
Bayesian approach is also better suited to complex and disparate data.
Incorporating various sources of information into mechanistic models can
be an immensely difficult due to the increased complexity involved. Only
thanks to tandem advances in increasing computational power and
hierarchical statistical methodology have we been able to tackle such
intuitively important complexity (and the potentially new available data
that accompanies it) such as spatial distribution, heterogeneities of
space, time, and individuals, to shift to ecosystem-based approaches
from single-species based approaches. Without the need to formulate
mechanisms, many modern machine learning algorithms can leverage
potential information from all available sources of data directly. The
algorithms can recognize unanticipated or subtle patterns in large data
sets that enable more accurate predictions than mechanistic models that
are formulated at a more macroscopic level.

@@

 \textbf{\emph{Possible additional discussion material?}}

\begin{itemize}
\item
  Discuss constant escapement in model, in policies.
\item
  Limitations of this comparison: Are the maximum-likelihood solutions a
  straw man?
\item
  Discussion of alternative related approaches: POMDP/MOMDP
\item
  Online learning
\item
  Multiple step-ahead predictions
\item
  Explicitly accomidating additional uncertainties
\item
  Improving inference of optimal policy from the GP 
\end{itemize}

@@

\subsubsection{The Gaussian process}

@@

\begin{longtable}[c]{@{}ll@{}}
\hline\noalign{\medskip}
parameter & interpretation
\\\noalign{\medskip}
\hline\noalign{\medskip}
$\sigma^2$ & The process noise (from the kernel)
\\\noalign{\medskip}
$\tau^2$ & Variance around the mean
\\\noalign{\medskip}
$\beta_0$ & The mean is given by a linear model of slope $\beta$
\\\noalign{\medskip}
$d$ & The length-scale of the covariance function
\\\noalign{\medskip}
$n$ & The observation error
\\\noalign{\medskip}
\hline
\noalign{\medskip}
\caption{Table of parameters for the Gaussian process}
\end{longtable}

@@

The posterior Gaussian process is completely defined by the expected
value and covariance matrix at a defined set of training points. For
simplicity we will consider a these points to fall on a discrete,
uniform grid $x$ of 101 points from 0 to 15 (1.5 times the positive
equilibrium $K$). Again to keep things simple we will use this same grid
discretization for the parametric approach. Other options for choosing
the grid points, including collocation methods and functional basis
expansion (or even using Gaussian processes in place of the discrete
optimization; an entirely different context in which GP can be used in
SDP, see (Deisenroth, Rasmussen, and Peters 2009)) could also be
considered.

@@

\textbf{\emph{Can we get this figure into the main text with appropriate
discussion without getting too technical?}}

\begin{figure}[htbp]
\centering
\includegraphics{figure/posteriors.png}
\caption{Histogram of posterior distributions for the estimated Gaussian
Process shown in Figure 1. Prior distributions overlaid.}
\end{figure}

@@

Bestelmeyer, Brandon T., Michael C. Duniway, Darren K. James, Laura M.
Burkett, and Kris M. Havstad. 2012. ``A test of critical thresholds and
their indicators in a desertification-prone ecosystem: more resilience
than we thought.'' Ed. Katharine Suding. \emph{Ecology Letters} (dec).
doi:10.1111/ele.12045. \url{http://doi.wiley.com/10.1111/ele.12045}.

@@
